---
title: "In-class Exercise 2"
author: "Amelia Chua"
execute: 
  warning: false
  message: false
number-sections: true
format: html
editor: visual
---

## Getting Started

In this exercise, we will be using the following packages:

-   sf: for importing, managing, and processing geospatial data

-   tmap: for performing data science tasks such as importing, wrangling and visualising data

-   tidyverse: for performing data science tasks such as importing, wrangling and visualising data

-   spdep: used to create spatial weights matrix objects from polygon 'contiguities'

-   funModeling: used for rapid Exploratory Data Analysis

We will use *pacman,* a R package management tool to install and load the packages.

```{r}
pacman::p_load(sf, tmap, tidyverse, spdep, funModeling)
```

## Importing Geospatial Data

In this in-class exercise, two geospatial data sets will be used:

-   Nigeria Level-2 Administrative Boundary (also known as Local Government Area or LGA) polygon feature GIS data. The data was obtained from geoBoundaries.

-   WPdx+ data set that was obtained from Water Point Data Exchange (WPdx). It consists of water point related data from rural areas at the water point or small water scheme level. The entire set of data includes countries other than Nigeria. Hence, we will be performing data pre-processing to extract the relevant data. This is also in the shape file format.

### Import Nigeria LGA boundary data

```{r}
nga <- st_read(dsn = "data\\geospatial\\",
                   layer = "geodata_NGA_ADM2",
                   crs = 4326)
```

### Import Nigeria Water Points data

The code chunk below will import our geospatial data. We indicated crs = 4326 as we know that this data is in the wgs84 format. We know this by looking at the projection file (.prj contains projection information).

While there are various file formats available for download e.g. .csv, .shp. The benefit of using the .shp file is that we can save one step of converting the .csv file into geospatial file using the latitude and longitude data.

```{r}
#| eval: false
wp <- st_read(dsn = "data\\geospatial\\",
                   layer = "wp_geodata",
                   crs = 4326) %>% 
  filter(clean_coun == "Nigeria")
```

Next, we will use `write_rds()` of **readr** package to save the extracted sf data table (i.e. wp) into an output file in .rds data format. The output file is called *wp_nga.rds* and it is saved in *geospatial* sub-folder.

```{r}
#| eval: false
wp_nga <- write_rds(wp, 
                   "data\\geospatial\\wp_nga.rds") 
```

## Data Wrangling

### Recoding NGA values into string

We will read the saved .rsd file. It will look at the `status_cle` column and replace NA values with 'unknown' using *replace_na().*

```{r}
#| eval: false
wp_nga <- read_rds('data\\geospatial\\wp_nga.rds') %>%
  mutate(status_cle = replace_na(status_cle, "Unknown"))
```

### Exploratory Data Analysis

In the code chunk below, *freq()* of **funModelling** package is used to display the distribution of `status_cle` field in `wp_nga`.

```{r}
#| eval: false
freq(data = wp_nga,
      input = 'status_cle')
```

From the output above, we can see that there are 9 categories. Almost half of all water pipes are functional and approximately 11% with status unknown.

### Extract functional water points

In the code chunk below, *filter()* of **dplyr** package is used to select functional water points.

```{r}
#| eval: false
wpt_functional <- wp_nga %>%
  filter(status_cle %in%
           c("Functional",
             "Functional but not in use",
             "Functional but needs repair"))
```

Using the below code chunk, we can see the breakdown of functional statuses.

```{r}
#| eval: false
freq(data = wpt_functional, input = 'status_cle')
```

Close to 80% of the water points are fully functional. The rest requires repair or are not in use.

### Extract non-functional water points

We can do the same for non-functional water points.

```{r}
#| eval: false
wpt_nonfunctional <- wp_nga %>%
  filter(status_cle %in%
           c("Non-Functional",
             "Non-Functional due to dry season",
             "Non Functional due to dry season",
             "Abandoned/Decommissioned",
             "Abandoned"))
```

Using the below code chunk, we can see the breakdown of non-functional statuses.

```{r}
#| eval: false
freq(data = wpt_nonfunctional, input = 'status_cle')
```

There is 90% of water points that are non-functional, another 7% are non-functional due to the weather. The remaining are abandoned or decommissioned.

### Extract water point with unknown class

```{r}
#| eval: false
wpt_unknown <- wp_nga %>%
  filter(status_cle == "Unknown")
```

### Performing Point-in-Polygon count

In the code chunk below, we will use *st_intersect()* yo identify which LGA respectively types of water points are located in and *lengths()* to count the number of water points that fall inside each LGA area.

```{r}
#| eval: false
nga_wp <- nga %>%
  mutate(`total_wpt` = lengths(
    st_intersects(nga, wp_nga))) %>%
  mutate(`wpt_functional` = lengths(
    st_intersects(nga, wpt_functional))) %>%
  mutate(`wpt_nonfunctional` = lengths(
    st_intersects(nga, wpt_nonfunctional))) %>%
  mutate(`wpt_unknown` = lengths(
    st_intersects(nga, wpt_unknown)))
```

## Saving the Analytical Data Table

```{r}
#| eval: false
nga_wp <- nga_wp %>%
  mutate(pct_functional = `wpt_functional`/ `total_wpt`) %>%
  mutate(pct_nonfunctional = `wpt_nonfunctional`/ `total_wpt`) %>%
  select(1, 6:10)
```

Things to learn from the code chunk above:

-   `mutate()` of **dplyr** package is used to derive two fields namely *pct_functional* and *pct_non-functional*.

-   to keep the file size small, `select()` of **dplyr** is used to retain selected fields

Let's save the new sf data table into .rds format.

```{r}
#| eval: false
write_rds(nga_wp,
          "data\\geospatial\\wp_nga.rds")
```

## Visualising the Spatial Distribution of Water Points

In the code chuck below, we will use *qtm()* to quickly plot the spatial distribution of water points of different categories. We will use *tmap_arrange()* to show the plots together.

```{r}
#| echo: false
nga_wp <- read_rds('data\\geospatial\\wp_nga.rds')
```

```{r}
total <- qtm(nga_wp, fill = 'total_wpt')
func <- qtm(nga_wp, fill = 'wpt_functional')
nonfunc <- qtm(nga_wp, fill = 'wpt_nonfunctional')
unknown <- qtm(nga_wp, fill = 'wpt_unknown')

tmap_arrange(total, unknown, func, nonfunc, asp = 1, ncol = 2)
```

From the maps above, we can see that there are some areas with little (\<200) water points and areas with \>800 water points. Also, towards the southern area of the map, we can see that there are areas where water pumps statuses are unknown, probably due to lack of updating. It appears that the top center of Nigeria have higher amount of functional water points.

## Conclusion

From what we have achieved in this exercise, we have grounds to believe that there might exist some relationship in how water points co-occur in Nigeria. We will explore more in Take-Home Exercise 1.
