---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview 

In this hands-on exercise, I learned how to import, and wrangle geospatial data using appropriate R packages.

## Getting Started

pacman is a R package management tool. It provides intuitively named functions for the base functions. The code chunk below installs and loads sf and tidyverse packages into R environment:

```{r}
pacman::p_load(sf, tidyverse)

```

An alternate way to install and import the libraries is as follows:

```{r}
packages = c('sf','tidyverse')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

## Import Geospatial Data

In this section, I learned how to import the following geospatial data into R by using [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package:

-   `MP14_SUBZONE_WEB_PL`, a polygon feature layer in ESRI shapefile format,

-   `CyclingPath`, a line feature layer in ESRI shapefile format, and

-   `PreSchool`, a point feature layer in kml file format.

### Import polygon feature date in shapefile format (.shp)

The code chunk below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame.

When the input geospatial data is in shapefile format, two arguments will be used, namely: `dsn` to define the data path and `layer` to provide the shapefile name. Also note that no extension such as .shp, .dbf, .prj and .shx are needed.

```{r}
mpsz = st_read(dsn="data\\geospatial",
               layer = "MP14_SUBZONE_WEB_PL")

```

The output shows that there are 323 multipolygon features and 15 fields. The Bounding box provides the x extend and y extend of the data.
